Learning Causally-Generated Stationary Time Series

### 2017-02-24 doc/paper.tex
Updates

One of the focusses in this work is to develop flexible prior distributions over the filter $h$ that can support essentially arbitrarily complex structure. We therefore choose to model it using a Gaussian process $h \sim \mathcal{GP}(0,k_h(t,t'))$. The covariance function describing the prior over the filter$k_h(t,t')$ should be carefully chosen. One important constraint on the covariance function comes from the fact that every real-world signal $f$ has finite power. This can be satisfied by letting the filter variance decay to zero at infinity sufficiently quickly \citet{Tobar:2015:Learning_Stationary}: let $h\cond g = w g$ where $w(t)= \exp(- \alpha t^2)$ and $g \sim \mathcal{GP}(0,k_g(t-t'))$ has finite power; in that case, % RET there's a question as to whether exponential decay might be more appropriate in the current case
\begin{align*}
    \mathbb{V}[f(t)]
    &= \int_{0}\exp(-2 \alpha t^2)\mathbb{V}[g(\tau)]\id{\tau} \\
    &\le k_g(0) \int_0 \exp(- 2 \alpha t^2) \id{\tau} < \infty,
\end{align*}
which could be infinite otherwise.
Equivalently, we let $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}))k_g(t-t')$. To retain flexibility in the prior on $h$, we let $k_g$ be an exponentiated quadratic. We thus have that $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}) - \gamma(t-t')^2)$. Note that $\alpha$ determines the typical temporal extent of the filter and $\gamma$ the typical time-scale over which it varies. These assumptions also serve to make inference well-posed; without them, any shifted version of the filter results in an identical statistical model for the data meaning the model is unidentifiable.

The prior distributions over $x$ and $h$ induce a prior distribution on $f$. Further including a scale $\sigma_f$ to control $f$'s prior power, we call this prior on $f$ the Causal Gaussian Process Convolution Model (CGPCM). To recapitulate, the CGPCM is formulated as follows:

@@

\subsection{Learning Rough Signals}
Consider a Gaussian process with a fixed kernel. Then, according to \cref{mod:cgpcm2}'s interpretation, the CGPCM should be able to learn this kernel. We compare the CGPCM and the GPCM.

@@

\Cref{fig:toy_acausal_sample} shows the results for fitting the GPCM and CGPCM to a noisy random sample from the GPCM. The sample consists of 300 data points. We used 51 evenly-spaced inducing points for $h$ and 80 evenly-spaced inducing point for $x$. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed.

@@

\Cref{fig:toy_causal_sample} shows the results for fitting the GPCM and CGPCM to instead a noiseless random sample from the CGPCM. Since the drawn filter $h$ satisfies $|h(0)|>0$, the sample is rough and thus appears noisy. The challenge for the models is to not mistake this sample roughness for observation noise.

Observe that the CGPCM fits the filter $h$ and kernel $k_{f \cond h}$ well, whereas GPCM's fit is more mediocre. Despite roughly getting the correlation structure right, the GPCM mistakes the sample roughness for observation noisy and provides a poor fit of the function $f\cond h$. The CGPCM, on the other hand, correctly recognises the sample roughness and provides an excellent fit of $f \cond h$. This discrepancy is explained by inspecting the fit of the PSD: the GPCM is unable to fit the higher frequencies in the sample, whereas the CGPCM does so succesfully.

@@

\section{Discussion}
We have presented the CGPCM as a model for causally-generated stationary time series. Inference is performed by finding an initial approximation through a variational MF approach, and then refining this approximation through a variational SMF approach combined with ESS. The GPCM has been validated on synthesis and real-world signals. {\color{red} some here}

>>>>

One of the focusses in this work is to develop flexible prior distributions over the filter $h$ that can support essentially arbitrarily complex structure. We therefore choose to model it using a Gaussian process $h \sim \mathcal{GP}(0,k_h(t,t'))$. The covariance function describing the prior over the filter$k_h(t,t')$ should be carefully chosen. One important constraint on the covariance function comes from the fact that every real-world signal $f$ has finite power. This can be satisfied by letting the filter variance decay to zero at infinity sufficiently quickly \citet{Tobar:2015:Learning_Stationary}: let $h\cond g = w g$ where $w(t)= \exp(- \alpha t^2)$ and $g \sim \mathcal{GP}(0,k_g(t-t'))$ has finite power; in that case, % RET there's a question as to whether exponential decay might be more appropriate in the current case
\begin{align*}
    \mathbb{V}[f(t)]
    &= \int_{0}\exp(-2 \alpha t^2)\mathbb{V}[g(\tau)]\id{\tau} \\
    &\le k_g(0) \int_0 \exp(- 2 \alpha t^2) \id{\tau} < \infty,
\end{align*}
which could be infinite otherwise.
Equivalently, we let $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}))k_g(t-t')$. To retain flexibility in the prior on $h$, we let $k_g$ be an exponentiated quadratic. We thus have that $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}) - \gamma(t-t')^2)$. Note that $\alpha$ determines $k_f$'s length scale and $\gamma$ its flexibility.

Through specifying a prior on $x$ and a prior on $h$ we have specified a prior on $f$. Further including a scale $\sigma_f$ to control $f$'s prior power, we call this prior on $f$ the Causal Gaussian Process Convolution Model (CGPCM). To recapitulate, the CGPCM is formulated as follows:

@@

\subsection{Learning Rough Signals}
In \cref{subsec:roughness} we showed that the CGPCM models both smooth and rough signals, whereas the GPCM only models smooth signals. We perform two experiments that show the significance of this modelling capability.

@@

\Cref{fig:toy_acausal_sample} shows the results for fitting the GPCM and CGPCM to a noisy random sample from the GPCM. The sample consists of 300 data points. We used 51 evenly-spaced inducing points for $h$ and 80 evenly-spaced inducing points for $x$. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed.

@@

\Cref{fig:toy_causal_sample} shows the results for fitting the GPCM and CGPCM to instead a \textit{noiseless} random sample from the CGPCM. Since the drawn filter $h$ satisfies $|h(0)|>0$, the sample is rough and thus appears noisy. The challenge for the models is to not mistake this sample roughness for observation noise.

Observe that the CGPCM fits the filter $h$ and kernel $k_{f \cond h}$ well, whereas GPCM's fit is more mediocre. Despite roughly getting the correlation structure right, the GPCM mistakes the sample roughness for observation noise and provides a poor fit of the function $f\cond h$. The CGPCM, on the other hand, correctly recognises the sample roughness and provides an excellent fit of $f \cond h$. This discrepancy is explained by inspecting the fit of the PSD: the GPCM is unable to fit the higher frequencies in the sample, whereas the CGPCM does so succesfully.

@@

\section{Discussion}
We have presented the CGPCM as a model for causally-generated stationary time series. Inference in the CGPCM is performed by finding an initial approximation through a variational MF approach, and then refining this approximation through a variational SMF approach combined with ESS. The GPCM has been validated on synthesis and real-world signals. {\color{red} some here}
